---
title: "P8106 Data Science II - Midterm Project"
author: "William Donovan (wd2328), Hantang Qin (hq2229), Yongyan Liu (yl6107)"
date: "2026-03-23"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(caret)
library(glmnet)
library(earth)
library(mgcv)
library(pls)
library(splines)
library(pdp)
library(vip)
library(corrplot)
library(knitr)
library(gridExtra)
```

# Data Preparation

```{r data-load}
load("datA.RData")
load("datB.RData")

# Remove ID, convert categorical variables to factors
process_data <- function(df) {
  df %>%
    select(-id) %>%
    mutate(
      menopause = factor(menopause),
      race = factor(race),
      tumor_grade = factor(tumor_grade),
      ER = factor(ER),
      PR = factor(PR),
      HER2 = factor(HER2),
      stage = factor(stage),
      surgery = factor(surgery),
      chemo = factor(chemo),
      hormonal_therapy = factor(hormonal_therapy)
    )
}

A <- process_data(datA)
B <- process_data(datB)

# Design matrix for methods requiring matrix input
x_A <- model.matrix(log_PFS ~ ., data = A)[, -1]
y_A <- A$log_PFS
x_B <- model.matrix(log_PFS ~ ., data = B)[, -1]
y_B <- B$log_PFS
```

# Part I: Prediction Model for PFS (1990–1995 Cohort)

## Exploratory Data Analysis

```{r eda-feature, fig.cap="Figure 1: Scatter plots of continuous predictors vs. log_PFS.", fig.width=10, fig.height=8}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(x = A %>% select(age, bmi, tumor_size, lymph_nodes, Ki67),
            y = A$log_PFS,
            plot = "scatter",
            span = .5,
            type = c("p", "smooth"),
            layout = c(3, 2))
```

```{r eda-boxplot, fig.cap="Figure 2: Boxplots of log_PFS by categorical predictors.", fig.width=12, fig.height=8}
cat_vars <- c("menopause", "race", "tumor_grade", "ER", "PR", "HER2",
              "stage", "surgery", "chemo", "hormonal_therapy")

A %>%
  select(all_of(cat_vars), log_PFS) %>%
  pivot_longer(-log_PFS, names_to = "variable", values_to = "level") %>%
  ggplot(aes(x = factor(level), y = log_PFS)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  facet_wrap(~ variable, scales = "free_x", nrow = 2) +
  labs(x = "Level", y = "Log PFS") +
  theme_bw()
```

```{r eda-corr, fig.cap="Figure 3: Correlation matrix of continuous variables.", fig.width=6, fig.height=5}
cor_mat <- cor(A %>% select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS))
corrplot(cor_mat, method = "circle", type = "full", addCoef.col = "black",
         tl.col = "black", number.cex = 0.7)
```

```{r eda-response, fig.cap="Figure 4: Distribution of log-transformed PFS.", fig.width=6, fig.height=4}
ggplot(A, aes(x = log_PFS)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 0.8) +
  labs(x = "Log-transformed PFS (months)", y = "Density") +
  theme_bw()
```

```{r summary-table-cont}
# Table 1a: Continuous variable summary
cont_summary <- A %>%
  select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(Mean = round(mean(Value), 2),
            SD = round(sd(Value), 2),
            Median = round(median(Value), 2),
            Min = round(min(Value), 2),
            Max = round(max(Value), 2))
kable(cont_summary, caption = "Table 1a: Summary of continuous variables (datA).")
```

```{r summary-table-cat}
# Table 1b: Categorical variable summary
cat_summary <- A %>%
  select(all_of(cat_vars)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Level") %>%
  count(Variable, Level) %>%
  group_by(Variable) %>%
  mutate(Percentage = round(100 * n / sum(n), 1),
         Summary = paste0(n, " (", Percentage, "%)")) %>%
  select(Variable, Level, Summary)
kable(cat_summary, caption = "Table 1b: Summary of categorical variables (datA).")
```

## Model Training

We use 10-fold cross-validation to train and compare four candidate models: elastic net, partial least squares (PLS), multivariate adaptive regression splines (MARS), and generalized additive model (GAM). To ensure a fair comparison, the same CV fold indices are used across all models.

```{r cv-setup}
set.seed(8106)
cv_folds <- createFolds(y_A, k = 10, returnTrain = TRUE)
ctrl <- trainControl(method = "cv", number = 10, index = cv_folds)
```

### Elastic Net

```{r enet}
set.seed(8106)
enet_grid <- expand.grid(
  alpha = seq(0, 1, length = 11),
  lambda = exp(seq(2, -6, length = 100))
)

enet_fit <- train(x_A, y_A,
                  method = "glmnet",
                  tuneGrid = enet_grid,
                  trControl = ctrl)

enet_fit$bestTune
```

```{r enet-plot, fig.cap="Figure 5: Elastic net CV RMSE.", fig.width=8, fig.height=5}
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))
plot(enet_fit, par.settings = myPar, xTrans = log)
```

### PLS

```{r pls}
set.seed(8106)
pls_fit <- train(x_A, y_A,
                 method = "pls",
                 tuneGrid = data.frame(ncomp = 1:15),
                 trControl = ctrl,
                 preProcess = c("center", "scale"))
pls_fit$bestTune
```

```{r pls-plot, fig.cap="Figure 6: PLS CV RMSE by number of components.", fig.width=6, fig.height=4}
plot(pls_fit)
```

### MARS

```{r mars}
set.seed(8106)
mars_grid <- expand.grid(degree = 1:3, nprune = 2:18)

mars_fit <- train(x_A, y_A,
                  method = "earth",
                  tuneGrid = mars_grid,
                  trControl = ctrl)

mars_fit$bestTune
```

```{r mars-plot, fig.cap="Figure 7: MARS CV RMSE across tuning parameters.", fig.width=7, fig.height=5}
plot(mars_fit)
```

### GAM

We fit GAM using `mgcv::gam()` directly with smooth terms applied only to continuous predictors, preserving categorical variables as linear terms.

```{r gam}
set.seed(8106)

# Fit GAM with proper formula: s() for continuous, linear for factors
gam_formula <- log_PFS ~ s(age) + menopause + race + s(bmi) +
  s(tumor_size) + tumor_grade + s(lymph_nodes) + ER + PR + HER2 +
  s(Ki67) + stage + surgery + chemo + hormonal_therapy

# Manual 10-fold CV for GAM to use the same folds
gam_cv_results <- sapply(seq_along(cv_folds), function(i) {
  train_idx <- cv_folds[[i]]
  gam_mod <- gam(gam_formula, data = A[train_idx, ])
  pred <- predict(gam_mod, newdata = A[-train_idx, ])
  RMSE(pred, y_A[-train_idx])
})

cat("GAM 10-fold CV RMSE:", round(mean(gam_cv_results), 4), "\n")

# Fit final GAM on full data
gam_fit_full <- gam(gam_formula, data = A)
```

### Model Comparison

```{r resamp, fig.cap="Figure 8: Cross-validated RMSE comparison of candidate models.", fig.width=7, fig.height=4}
resamp <- resamples(list(
  ElasticNet = enet_fit,
  PLS = pls_fit,
  MARS = mars_fit
))

summary(resamp)

# Add GAM results manually for display
rmse_caret <- resamp$values %>%
  select(ends_with("~RMSE")) %>%
  colMeans(na.rm = TRUE)
names(rmse_caret) <- gsub("~RMSE", "", names(rmse_caret))
rmse_all <- c(rmse_caret, GAM = mean(gam_cv_results))

r2_caret <- resamp$values %>%
  select(ends_with("~Rsquared")) %>%
  colMeans(na.rm = TRUE)
names(r2_caret) <- gsub("~Rsquared", "", names(r2_caret))

# GAM R2 via CV
gam_r2_results <- sapply(seq_along(cv_folds), function(i) {
  train_idx <- cv_folds[[i]]
  gam_mod <- gam(gam_formula, data = A[train_idx, ])
  pred <- predict(gam_mod, newdata = A[-train_idx, ])
  caret::R2(pred, y_A[-train_idx])
})
r2_all <- c(r2_caret, GAM = mean(gam_r2_results))

comparison_table <- data.frame(
  Model = names(rmse_all),
  CV_RMSE = round(rmse_all, 4),
  CV_R2 = round(r2_all, 4)
)
kable(comparison_table, row.names = FALSE,
      caption = "Table 2: Cross-validated performance of candidate models (datA).")

bwplot(resamp, metric = "RMSE")
```

```{r best-model}
best_name <- names(which.min(rmse_all))
cat("Best model by CV RMSE:", best_name, "=", round(min(rmse_all), 4), "\n")
```

## Model Interpretation

### Elastic Net Coefficients

```{r enet-coef}
coef_enet <- coef(enet_fit$finalModel, s = enet_fit$bestTune$lambda)
coef_df <- data.frame(
  Variable = rownames(coef_enet),
  Coefficient = round(as.numeric(coef_enet), 4)
) %>% filter(Coefficient != 0) %>% arrange(desc(abs(Coefficient)))

kable(coef_df, caption = "Table 3: Non-zero elastic net coefficients.",
      row.names = FALSE)
```

### MARS Model Summary

```{r mars-summary}
summary(mars_fit$finalModel)
```

```{r mars-coef}
mars_coef <- data.frame(
  Term = names(coef(mars_fit$finalModel)),
  Coefficient = round(as.numeric(coef(mars_fit$finalModel)), 4)
)
kable(mars_coef, caption = "Table 4: MARS basis function coefficients.",
      row.names = FALSE)
```

### GAM Summary

```{r gam-summary}
summary(gam_fit_full)
```

```{r gam-plot, fig.cap="Figure 9: GAM smooth function estimates.", fig.width=10, fig.height=8}
par(mfrow = c(2, 3))
plot(gam_fit_full, shade = TRUE)
par(mfrow = c(1, 1))
```

### Variable Importance

```{r vip-all, fig.cap="Figure 10: Variable importance from MARS and elastic net.", fig.width=10, fig.height=5}
p1 <- vip(mars_fit, num_features = 10) + theme_bw() + ggtitle("MARS")
p2 <- vip(enet_fit, num_features = 10) + theme_bw() + ggtitle("Elastic Net")
grid.arrange(p1, p2, ncol = 2)
```

### Partial Dependence Plots

```{r pdp-mars, fig.cap="Figure 11: Partial dependence plots for top predictors (MARS).", fig.width=10, fig.height=8}
top_vars <- vip(mars_fit, num_features = 6)$data$Variable

pdp_list <- lapply(top_vars, function(v) {
  pd <- partial(mars_fit, pred.var = v, grid.resolution = 50)
  autoplot(pd) + theme_bw() + ylab("log_PFS")
})
grid.arrange(grobs = pdp_list, ncol = 3)
```

### Residual Diagnostics (Training Set)

```{r resid-train, fig.cap="Figure 12: Residual diagnostics for MARS on training data.", fig.width=8, fig.height=4}
pred_train <- as.vector(predict(mars_fit, newdata = x_A))
resid_train <- y_A - pred_train

p1 <- ggplot(data.frame(Predicted = pred_train, Residual = resid_train),
             aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted", y = "Residual", title = "Residuals vs Fitted") +
  theme_bw()

p2 <- ggplot(data.frame(Residual = resid_train), aes(sample = Residual)) +
  stat_qq(alpha = 0.2, size = 0.8) +
  stat_qq_line(color = "red") +
  labs(title = "Normal Q-Q Plot") +
  theme_bw()

grid.arrange(p1, p2, ncol = 2)
```

---

# Part II: Generalization to the 2000–2005 Cohort

## Apply Model A to Data B

```{r gen-eval}
pred_enet_B <- predict(enet_fit, newdata = x_B)
pred_pls_B  <- predict(pls_fit,  newdata = x_B)
pred_mars_B <- predict(mars_fit, newdata = x_B)
pred_gam_B  <- predict(gam_fit_full, newdata = B)

gen_table <- data.frame(
  Model = c("ElasticNet", "PLS", "MARS", "GAM"),
  CV_RMSE_A = round(rmse_all, 4),
  RMSE_on_B = round(c(RMSE(pred_enet_B, y_B),
                       RMSE(pred_pls_B, y_B),
                       RMSE(pred_mars_B, y_B),
                       RMSE(pred_gam_B, y_B)), 4),
  R2_on_B = round(c(caret::R2(pred_enet_B, y_B),
                     caret::R2(pred_pls_B, y_B),
                     caret::R2(pred_mars_B, y_B),
                     caret::R2(pred_gam_B, y_B)), 4)
)
gen_table$RMSE_increase <- paste0(round(100 * (gen_table$RMSE_on_B - gen_table$CV_RMSE_A) /
                                          gen_table$CV_RMSE_A, 1), "%")
kable(gen_table, caption = "Table 5: Generalization of Model A to datB.", row.names = FALSE)
```

```{r gen-plot, fig.cap="Figure 13: Observed vs. predicted log_PFS on datB (MARS from cohort A).", fig.width=6, fig.height=5}
pred_best_B <- as.vector(predict(mars_fit, newdata = x_B))

ggplot(data.frame(Observed = y_B, Predicted = pred_best_B),
       aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 0.8) +
  labs(x = "Observed log_PFS", y = "Predicted log_PFS") +
  theme_bw()
```

```{r resid-B, fig.cap="Figure 14: Residual distribution of Model A on datB.", fig.width=6, fig.height=4}
resid_B <- y_B - pred_best_B
ggplot(data.frame(Residual = resid_B), aes(x = Residual)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 0.8) +
  geom_vline(xintercept = mean(resid_B), color = "red", linetype = "dashed") +
  labs(x = "Residual", y = "Density",
       subtitle = paste("Mean residual =", round(mean(resid_B), 3))) +
  theme_bw()
```

## Distribution Comparison Between Cohorts

```{r dist-compare, fig.cap="Figure 15: Distribution comparison of continuous variables between cohorts.", fig.width=10, fig.height=8}
combined <- bind_rows(
  datA %>% mutate(Cohort = "1990-1995"),
  datB %>% mutate(Cohort = "2000-2005")
)

combined %>%
  select(log_PFS, tumor_size, lymph_nodes, Ki67, age, bmi, Cohort) %>%
  pivot_longer(-Cohort, names_to = "Variable", values_to = "Value") %>%
  ggplot(aes(x = Value, fill = Cohort)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ Variable, scales = "free", nrow = 2) +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")
```

```{r cat-compare, fig.cap="Figure 16: Categorical variable distributions across cohorts.", fig.width=10, fig.height=6}
combined %>%
  select(all_of(cat_vars), Cohort) %>%
  pivot_longer(-Cohort, names_to = "Variable", values_to = "Level") %>%
  count(Cohort, Variable, Level) %>%
  group_by(Cohort, Variable) %>%
  mutate(Prop = n / sum(n)) %>%
  ggplot(aes(x = factor(Level), y = Prop, fill = Cohort)) +
  geom_col(position = "dodge", alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free", nrow = 2) +
  labs(x = "Level", y = "Proportion") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")
```

```{r cohort-summary}
comp_A <- A %>%
  select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS) %>%
  summarise(across(everything(), ~ paste0(round(mean(.x), 2), " (", round(sd(.x), 2), ")"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Cohort A (1990-1995)")

comp_B <- B %>%
  select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS) %>%
  summarise(across(everything(), ~ paste0(round(mean(.x), 2), " (", round(sd(.x), 2), ")"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Cohort B (2000-2005)")

kable(left_join(comp_A, comp_B, by = "Variable"),
      caption = "Table 6: Mean (SD) comparison of continuous variables between cohorts.")
```

## Build New Model for datB

```{r model-B-setup}
set.seed(8106)
cv_folds_B <- createFolds(y_B, k = 10, returnTrain = TRUE)
ctrl_B <- trainControl(method = "cv", number = 10, index = cv_folds_B)
```

```{r model-B}
set.seed(8106)
enet_fit_B <- train(x_B, y_B, method = "glmnet",
                    tuneGrid = enet_grid, trControl = ctrl_B)

pls_fit_B <- train(x_B, y_B, method = "pls",
                   tuneGrid = data.frame(ncomp = 1:15),
                   trControl = ctrl_B, preProcess = c("center", "scale"))

mars_fit_B <- train(x_B, y_B, method = "earth",
                    tuneGrid = mars_grid, trControl = ctrl_B)

# GAM with proper formula on datB
gam_cv_B <- sapply(seq_along(cv_folds_B), function(i) {
  train_idx <- cv_folds_B[[i]]
  gam_mod <- gam(gam_formula, data = B[train_idx, ])
  pred <- predict(gam_mod, newdata = B[-train_idx, ])
  RMSE(pred, y_B[-train_idx])
})
gam_fit_B_full <- gam(gam_formula, data = B)

gam_r2_B <- sapply(seq_along(cv_folds_B), function(i) {
  train_idx <- cv_folds_B[[i]]
  gam_mod <- gam(gam_formula, data = B[train_idx, ])
  pred <- predict(gam_mod, newdata = B[-train_idx, ])
  caret::R2(pred, y_B[-train_idx])
})
```

```{r resamp-B, fig.cap="Figure 17: CV RMSE comparison for models trained on datB.", fig.width=7, fig.height=4}
resamp_B <- resamples(list(
  ElasticNet = enet_fit_B,
  PLS = pls_fit_B,
  MARS = mars_fit_B
))

rmse_caret_B <- resamp_B$values %>%
  select(ends_with("~RMSE")) %>%
  colMeans(na.rm = TRUE)
names(rmse_caret_B) <- gsub("~RMSE", "", names(rmse_caret_B))
rmse_all_B <- c(rmse_caret_B, GAM = mean(gam_cv_B))

r2_caret_B <- resamp_B$values %>%
  select(ends_with("~Rsquared")) %>%
  colMeans(na.rm = TRUE)
names(r2_caret_B) <- gsub("~Rsquared", "", names(r2_caret_B))
r2_all_B <- c(r2_caret_B, GAM = mean(gam_r2_B))

comp_B_table <- data.frame(
  Model = names(rmse_all_B),
  CV_RMSE = round(rmse_all_B, 4),
  CV_R2 = round(r2_all_B, 4)
)
kable(comp_B_table, row.names = FALSE,
      caption = "Table 7: Cross-validated performance of models on datB.")

bwplot(resamp_B, metric = "RMSE")
```

### Variable Importance Comparison

```{r vip-compare, fig.cap="Figure 18: Variable importance comparison (MARS) between cohorts.", fig.width=10, fig.height=5}
p1 <- vip(mars_fit, num_features = 10) + theme_bw() + ggtitle("Cohort A (1990-1995)")
p2 <- vip(mars_fit_B, num_features = 10) + theme_bw() + ggtitle("Cohort B (2000-2005)")
grid.arrange(p1, p2, ncol = 2)
```

### Coefficient Comparison (Elastic Net)

```{r coef-compare}
coef_A <- coef(enet_fit$finalModel, s = enet_fit$bestTune$lambda)
coef_Bnew <- coef(enet_fit_B$finalModel, s = enet_fit_B$bestTune$lambda)

coef_comp <- data.frame(
  Variable = rownames(coef_A),
  Cohort_A = round(as.numeric(coef_A), 4),
  Cohort_B = round(as.numeric(coef_Bnew), 4)
) %>%
  mutate(Change = round(Cohort_B - Cohort_A, 4)) %>%
  filter(Cohort_A != 0 | Cohort_B != 0) %>%
  arrange(desc(abs(Change)))

kable(coef_comp, caption = "Table 8: Elastic net coefficient comparison between cohorts.",
      row.names = FALSE)
```

### MARS Comparison

```{r mars-B-summary}
summary(mars_fit_B$finalModel)

mars_coef_B <- data.frame(
  Term = names(coef(mars_fit_B$finalModel)),
  Coefficient = round(as.numeric(coef(mars_fit_B$finalModel)), 4)
)
kable(mars_coef_B, caption = "Table 9: MARS basis function coefficients (datB).",
      row.names = FALSE)
```

### PDP Comparison

```{r pdp-compare, fig.cap="Figure 19: PDP comparison for top predictors between cohorts.", fig.width=10, fig.height=4}
top3 <- vip(mars_fit, num_features = 3)$data$Variable

pdp_compare <- lapply(top3, function(v) {
  pd_a <- partial(mars_fit, pred.var = v, grid.resolution = 50) %>%
    mutate(Cohort = "A (1990-1995)")
  pd_b <- partial(mars_fit_B, pred.var = v, grid.resolution = 50) %>%
    mutate(Cohort = "B (2000-2005)")
  pd_both <- bind_rows(pd_a, pd_b)
  ggplot(pd_both, aes(x = .data[[v]], y = yhat, color = Cohort)) +
    geom_line(linewidth = 1) +
    labs(y = "log_PFS", title = v) +
    theme_bw() + scale_color_brewer(palette = "Set1")
})
grid.arrange(grobs = pdp_compare, ncol = 3)
```

## Summary

```{r final-summary}
best_B_name <- names(which.min(rmse_all_B))

final_table <- data.frame(
  Metric = c("Best Model", "CV RMSE", "CV R-squared"),
  Cohort_A = c(best_name,
               round(min(rmse_all), 4),
               round(r2_all[best_name], 4)),
  Cohort_B = c(best_B_name,
               round(min(rmse_all_B), 4),
               round(r2_all_B[best_B_name], 4))
)
kable(final_table, caption = "Table 10: Best model summary for each cohort.", row.names = FALSE)
```
