---
title: "P8106 Data Science II - Midterm Project"
author: "William Donovan (wd2328), Hantang Qin (hq2229), Yongyan Liu (yl6107)"
date: "2026-03-23"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.path = "figures/", dev = "png", dpi = 300)
library(tidyverse)
library(caret)
library(glmnet)
library(earth)
library(mgcv)
library(pls)
library(splines)
library(pdp)
library(vip)
library(corrplot)
library(knitr)
library(gridExtra)
```

# Data Preparation

```{r data-load}
load("datA.RData")
load("datB.RData")

# Remove ID, convert categorical variables to factors
process_data <- function(df) {
  df %>%
    select(-id) %>%
    mutate(
      menopause = factor(menopause),
      race = factor(race),
      tumor_grade = factor(tumor_grade),
      ER = factor(ER),
      PR = factor(PR),
      HER2 = factor(HER2),
      stage = factor(stage),
      surgery = factor(surgery),
      chemo = factor(chemo),
      hormonal_therapy = factor(hormonal_therapy)
    )
}

A <- process_data(datA)
B <- process_data(datB)

# 80/20 train-test split for datA
set.seed(8106)
trainIndex_A <- createDataPartition(A$log_PFS, p = 0.8, list = FALSE)
train_A <- A[trainIndex_A, ]
test_A <- A[-trainIndex_A, ]

x_train_A <- model.matrix(log_PFS ~ ., data = train_A)[, -1]
y_train_A <- train_A$log_PFS
x_test_A <- model.matrix(log_PFS ~ ., data = test_A)[, -1]
y_test_A <- test_A$log_PFS

# Full design matrices (for generalization to datB)
x_A <- model.matrix(log_PFS ~ ., data = A)[, -1]
y_A <- A$log_PFS
x_B <- model.matrix(log_PFS ~ ., data = B)[, -1]
y_B <- B$log_PFS
```

# Part I: Prediction Model for PFS (1990--1995 Cohort)

## Exploratory Data Analysis

```{r eda-response, fig.cap="Distribution of log-transformed PFS (datA).", fig.width=5, fig.height=3}
ggplot(A, aes(x = log_PFS)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", alpha = 0.7) +
  geom_density(linewidth = 0.8) +
  labs(x = "Log-transformed PFS (months)", y = "Density") +
  theme_bw()
```

```{r eda-feature, fig.cap="Scatter plots of continuous predictors vs. log\\_PFS.", fig.width=7, fig.height=5}
theme1 <- trellis.par.get()
theme1$plot.symbol$col <- rgb(.2, .4, .2, .5)
theme1$plot.symbol$pch <- 16
theme1$plot.line$col <- rgb(.8, .1, .1, 1)
theme1$plot.line$lwd <- 2
theme1$strip.background$col <- rgb(.0, .2, .6, .2)
trellis.par.set(theme1)

featurePlot(x = A %>% select(age, bmi, tumor_size, lymph_nodes, Ki67),
            y = A$log_PFS,
            plot = "scatter",
            span = .5,
            type = c("p", "smooth"),
            layout = c(3, 2))
```

```{r eda-corr, fig.cap="Correlation matrix of continuous variables.", fig.width=5, fig.height=4}
cor_mat <- cor(A %>% select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS))
corrplot(cor_mat, method = "circle", type = "full", addCoef.col = "black",
         tl.col = "black", number.cex = 0.7)
```

```{r eda-boxplot, fig.cap="Boxplots of log\\_PFS by categorical predictors.", fig.width=7, fig.height=5}
cat_vars <- c("menopause", "race", "tumor_grade", "ER", "PR", "HER2",
              "stage", "surgery", "chemo", "hormonal_therapy")

A %>%
  select(all_of(cat_vars), log_PFS) %>%
  pivot_longer(-log_PFS, names_to = "variable", values_to = "level") %>%
  ggplot(aes(x = factor(level), y = log_PFS)) +
  geom_boxplot(fill = "steelblue", alpha = 0.5) +
  facet_wrap(~ variable, scales = "free_x", nrow = 2) +
  labs(x = "Level", y = "Log PFS") +
  theme_bw()
```

```{r summary-table-cont}
cont_summary <- A %>%
  select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Value") %>%
  group_by(Variable) %>%
  summarise(Mean = round(mean(Value), 2),
            SD = round(sd(Value), 2),
            Median = round(median(Value), 2),
            Min = round(min(Value), 2),
            Max = round(max(Value), 2))
kable(cont_summary, caption = "Summary of continuous variables (datA).")
```

```{r summary-table-cat}
cat_summary <- A %>%
  select(all_of(cat_vars)) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Level") %>%
  count(Variable, Level) %>%
  group_by(Variable) %>%
  mutate(Percentage = round(100 * n / sum(n), 1),
         Summary = paste0(n, " (", Percentage, "%)")) %>%
  select(Variable, Level, Summary)
kable(cat_summary, caption = "Summary of categorical variables (datA).")
```

## Model Training

We split datA into 80% training and 20% test sets. We use 10-fold cross-validation on the training set to tune hyperparameters with two selection rules: "best" (minimum CV RMSE) and "oneSE" (simplest model within 1 SE of the minimum). All models are then evaluated on the held-out test set. The same CV fold indices are used across all models for a fair comparison.

```{r cv-setup}
set.seed(8106)
cv_folds <- createFolds(y_train_A, k = 10, returnTrain = TRUE)
ctrl_best <- trainControl(method = "cv", number = 10, index = cv_folds,
                          selectionFunction = "best")
ctrl_1se  <- trainControl(method = "cv", number = 10, index = cv_folds,
                          selectionFunction = "oneSE")

enet_grid <- expand.grid(
  alpha = seq(0, 1, length = 11),
  lambda = exp(seq(2, -6, length = 100))
)
mars_grid <- expand.grid(degree = 1:3, nprune = 2:18)
```

### Elastic Net

```{r enet}
set.seed(8106)
# Step 1: best on full alpha x lambda grid
enet_fit <- train(x_train_A, y_train_A, method = "glmnet",
                  tuneGrid = enet_grid, trControl = ctrl_best)

# Step 2: fix best alpha, apply 1SE along lambda only
enet_grid_1se <- expand.grid(alpha = enet_fit$bestTune$alpha,
                             lambda = exp(seq(2, -6, length = 100)))
enet_fit_1se <- train(x_train_A, y_train_A, method = "glmnet",
                      tuneGrid = enet_grid_1se, trControl = ctrl_1se)

cat("Best: alpha =", enet_fit$bestTune$alpha, ", lambda =", enet_fit$bestTune$lambda, "\n")
cat("1SE:  alpha =", enet_fit_1se$bestTune$alpha, ", lambda =", enet_fit_1se$bestTune$lambda, "\n")
```

```{r enet-plot, fig.cap="Elastic net CV RMSE.", fig.width=8, fig.height=5}
myCol <- rainbow(25)
myPar <- list(superpose.symbol = list(col = myCol),
              superpose.line = list(col = myCol))
plot(enet_fit, par.settings = myPar, xTrans = log)
```

### PLS

```{r pls}
set.seed(8106)
pls_fit     <- train(x_train_A, y_train_A, method = "pls",
                     tuneGrid = data.frame(ncomp = 1:15),
                     trControl = ctrl_best, preProcess = c("center", "scale"))
pls_fit_1se <- train(x_train_A, y_train_A, method = "pls",
                     tuneGrid = data.frame(ncomp = 1:15),
                     trControl = ctrl_1se, preProcess = c("center", "scale"))

cat("Best ncomp:", pls_fit$bestTune$ncomp, "\n")
cat("1SE ncomp: ", pls_fit_1se$bestTune$ncomp, "\n")
```

```{r pls-plot, fig.cap="PLS CV RMSE by number of components.", fig.width=6, fig.height=4}
plot(pls_fit)
```

### MARS

```{r mars}
set.seed(8106)
mars_fit     <- train(x_train_A, y_train_A, method = "earth",
                      tuneGrid = mars_grid, trControl = ctrl_best)
mars_fit_1se <- train(x_train_A, y_train_A, method = "earth",
                      tuneGrid = mars_grid, trControl = ctrl_1se)

cat("Best:", paste(mars_fit$bestTune, collapse = ", "), "\n")
cat("1SE: ", paste(mars_fit_1se$bestTune, collapse = ", "), "\n")
```

```{r mars-plot, fig.cap="MARS CV RMSE across tuning parameters.", fig.width=7, fig.height=5}
plot(mars_fit)
```

### GAM

We fit GAM using `mgcv::gam()` directly with smooth terms (`s()`) on the five continuous predictors and parametric terms for all categorical predictors. Since `caret`'s GAM wrapper applies smoothing to all columns including dummy variables, we perform manual 10-fold CV using the same fold indices.

```{r gam}
set.seed(8106)

gam_formula <- log_PFS ~ s(age) + menopause + race + s(bmi) +
  s(tumor_size) + tumor_grade + s(lymph_nodes) + ER + PR + HER2 +
  s(Ki67) + stage + surgery + chemo + hormonal_therapy

gam_cv_results <- sapply(seq_along(cv_folds), function(i) {
  train_idx <- cv_folds[[i]]
  gam_mod <- gam(gam_formula, data = train_A[train_idx, ])
  pred <- predict(gam_mod, newdata = train_A[-train_idx, ])
  RMSE(pred, y_train_A[-train_idx])
})

cat("GAM 10-fold CV RMSE:", round(mean(gam_cv_results), 4), "\n")

gam_fit_full <- gam(gam_formula, data = train_A)
```

### Model Comparison and Test Set Evaluation

```{r model-compare-A}
# Helper: count number of original predictors used (caret models only)
count_vars <- function(fit) {
  imp <- varImp(fit)$importance
  nz <- rownames(imp)[!is.na(imp[,1]) & imp[,1] > 0]
  length(unique(gsub("[0-9]+$", "", nz)))
}

# Caret models: best and 1SE
caret_models <- list(
  "ElasticNet (best)" = enet_fit,     "ElasticNet (1se)" = enet_fit_1se,
  "PLS (best)"        = pls_fit,      "PLS (1se)"        = pls_fit_1se,
  "MARS (best)"       = mars_fit,     "MARS (1se)"       = mars_fit_1se
)

cv_results <- sapply(caret_models, function(m) merge(m$bestTune, m$results)$RMSE)
test_preds <- lapply(caret_models, function(m) predict(m, newdata = x_test_A))
test_rmse_all <- sapply(test_preds, function(p) RMSE(p, y_test_A))
n_vars <- sapply(caret_models, count_vars)

# GAM (manual CV, single entry)
gam_test_rmse <- RMSE(predict(gam_fit_full, newdata = test_A), y_test_A)

comparison_table <- rbind(
  data.frame(Model = names(caret_models),
             CV_RMSE = round(cv_results, 4),
             Test_RMSE = round(test_rmse_all, 4),
             Num_Predictors = n_vars),
  data.frame(Model = "GAM",
             CV_RMSE = round(mean(gam_cv_results), 4),
             Test_RMSE = round(gam_test_rmse, 4),
             Num_Predictors = 15)
)
kable(comparison_table, row.names = FALSE,
      caption = "Model performance with best and 1SE selection rules (datA).")
```

```{r best-model}
# For downstream use
test_rmse <- c(
  ElasticNet = unname(test_rmse_all["ElasticNet (best)"]),
  PLS        = unname(test_rmse_all["PLS (best)"]),
  MARS       = unname(test_rmse_all["MARS (best)"]),
  GAM        = gam_test_rmse
)

# CV RMSE and R2 for the best models (for report tables)
resamp <- resamples(list(
  ElasticNet = enet_fit, PLS = pls_fit, MARS = mars_fit
))
rmse_all <- c(
  sapply(caret_models[c(1,3,5)], function(m) merge(m$bestTune, m$results)$RMSE),
  GAM = mean(gam_cv_results)
)
names(rmse_all) <- c("ElasticNet", "PLS", "MARS", "GAM")

r2_caret <- resamp$values %>%
  select(ends_with("~Rsquared")) %>%
  colMeans(na.rm = TRUE)
names(r2_caret) <- gsub("~Rsquared", "", names(r2_caret))

gam_r2_results <- sapply(seq_along(cv_folds), function(i) {
  train_idx <- cv_folds[[i]]
  gam_mod <- gam(gam_formula, data = train_A[train_idx, ])
  pred <- predict(gam_mod, newdata = train_A[-train_idx, ])
  caret::R2(pred, y_train_A[-train_idx])
})
r2_all <- c(r2_caret, GAM = mean(gam_r2_results))

cat("Selected final model: MARS\n")
best_name <- "MARS"
```

## Model Interpretation

### Elastic Net Coefficients

```{r enet-coef}
coef_enet <- coef(enet_fit$finalModel, s = enet_fit$bestTune$lambda)
coef_df <- data.frame(
  Variable = rownames(coef_enet),
  Coefficient = round(as.numeric(coef_enet), 4)
) %>% filter(Coefficient != 0) %>% arrange(desc(abs(Coefficient)))

kable(coef_df, caption = "Non-zero elastic net coefficients.",
      row.names = FALSE)
```

### MARS Model Summary

```{r mars-summary}
summary(mars_fit$finalModel)
```

### GAM Summary

```{r gam-summary}
summary(gam_fit_full)
```

```{r gam-plot, fig.cap="GAM smooth function estimates (datA).", fig.width=7, fig.height=5}
par(mfrow = c(2, 3))
plot(gam_fit_full, shade = TRUE)
par(mfrow = c(1, 1))
```

### Variable Importance

```{r vip-all, fig.cap="Variable importance from MARS, Lasso, and PLS (datA).", fig.width=9, fig.height=3.5}
p1 <- vip(mars_fit, num_features = 10) + theme_bw() + ggtitle("MARS")
p2 <- vip(enet_fit, num_features = 10) + theme_bw() + ggtitle("Lasso")
p3 <- vip(pls_fit, num_features = 10) + theme_bw() + ggtitle("PLS")
grid.arrange(p1, p2, p3, ncol = 3)
```

### Partial Dependence Plots

```{r pdp-mars, fig.cap="Partial dependence plots for top 6 predictors (MARS, datA).", fig.width=7, fig.height=5}
top_vars <- vip(mars_fit, num_features = 6)$data$Variable

pdp_list <- lapply(top_vars, function(v) {
  pd <- partial(mars_fit, pred.var = v, grid.resolution = 50)
  autoplot(pd) + theme_bw() + ylab("log_PFS")
})
grid.arrange(grobs = pdp_list, ncol = 3)
```

### Residual Diagnostics

```{r resid-diag, fig.cap="Residual diagnostics for MARS on training (top) and test (bottom) data.", fig.width=7, fig.height=6}
pred_train <- as.vector(predict(mars_fit, newdata = x_train_A))
resid_train <- y_train_A - pred_train
pred_test <- as.vector(predict(mars_fit, newdata = x_test_A))
resid_test <- y_test_A - pred_test

p1 <- ggplot(data.frame(Predicted = pred_train, Residual = resid_train),
             aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted", y = "Residual", title = "Training: Residuals vs Fitted") +
  theme_bw()

p2 <- ggplot(data.frame(Residual = resid_train), aes(sample = Residual)) +
  stat_qq(alpha = 0.2, size = 0.8) + stat_qq_line(color = "red") +
  labs(title = "Training: Q-Q Plot") + theme_bw()

p3 <- ggplot(data.frame(Predicted = pred_test, Residual = resid_test),
             aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_hline(yintercept = 0, color = "red") +
  labs(x = "Predicted", y = "Residual", title = "Test: Residuals vs Fitted") +
  theme_bw()

p4 <- ggplot(data.frame(Residual = resid_test), aes(sample = Residual)) +
  stat_qq(alpha = 0.3, size = 0.8) + stat_qq_line(color = "red") +
  labs(title = "Test: Q-Q Plot") + theme_bw()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

---

# Part II: Generalization to the 2000--2005 Cohort

## Apply Model A to Data B

```{r gen-eval}
# Predictions on datB for all models (best + 1SE + GAM)
gen_models <- list(
  "ElasticNet (best)" = enet_fit,     "ElasticNet (1se)" = enet_fit_1se,
  "PLS (best)"        = pls_fit,      "PLS (1se)"        = pls_fit_1se,
  "MARS (best)"       = mars_fit,     "MARS (1se)"       = mars_fit_1se
)

gen_preds <- lapply(gen_models, function(m) predict(m, newdata = x_B))
gen_rmse <- sapply(gen_preds, function(p) RMSE(p, y_B))

# GAM
gen_rmse_gam <- RMSE(predict(gam_fit_full, newdata = B), y_B)

# Test RMSE on A for reference
test_rmse_ref <- c(
  comparison_table$Test_RMSE[comparison_table$Model == "ElasticNet (best)"],
  comparison_table$Test_RMSE[comparison_table$Model == "ElasticNet (1se)"],
  comparison_table$Test_RMSE[comparison_table$Model == "PLS (best)"],
  comparison_table$Test_RMSE[comparison_table$Model == "PLS (1se)"],
  comparison_table$Test_RMSE[comparison_table$Model == "MARS (best)"],
  comparison_table$Test_RMSE[comparison_table$Model == "MARS (1se)"],
  comparison_table$Test_RMSE[comparison_table$Model == "GAM"]
)

gen_table <- data.frame(
  Model = c(names(gen_models), "GAM"),
  Test_RMSE_A = test_rmse_ref,
  RMSE_on_B = round(c(gen_rmse, gen_rmse_gam), 4)
)
gen_table$Pct_Increase <- paste0(round(100 * (gen_table$RMSE_on_B - gen_table$Test_RMSE_A) /
                                         gen_table$Test_RMSE_A, 1), "%")
kable(gen_table, row.names = FALSE,
      caption = "Generalization of Model A to datB.")
```

```{r gen-plot, fig.cap="Observed vs. predicted log\\_PFS on datB (MARS from cohort A).", fig.width=5, fig.height=4}
pred_best_B <- as.vector(predict(mars_fit, newdata = x_B))

ggplot(data.frame(Observed = y_B, Predicted = pred_best_B),
       aes(x = Observed, y = Predicted)) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 0.8) +
  labs(x = "Observed log_PFS", y = "Predicted log_PFS") +
  theme_bw()
```

```{r resid-B, fig.cap="Residual comparison: MARS on test set A vs. full datB.", fig.width=6, fig.height=4}
resid_test_A <- y_test_A - as.vector(predict(mars_fit, newdata = x_test_A))
resid_B <- y_B - pred_best_B

resid_df <- bind_rows(
  data.frame(Residual = resid_test_A, Dataset = paste0("Test A (mean = ", round(mean(resid_test_A), 3), ")")),
  data.frame(Residual = resid_B, Dataset = paste0("datB (mean = ", round(mean(resid_B), 3), ")"))
)

ggplot(resid_df, aes(x = Residual, fill = Dataset)) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, color = "black", linetype = "dashed") +
  labs(x = "Residual", y = "Density") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")
```

## Distribution Comparison Between Cohorts

```{r dist-compare, fig.cap="Distribution comparison of continuous variables between cohorts.", fig.width=7, fig.height=5}
combined <- bind_rows(
  datA %>% mutate(Cohort = "1990-1995"),
  datB %>% mutate(Cohort = "2000-2005")
)

p <- combined %>%
  select(log_PFS, age, bmi, tumor_size, lymph_nodes, Ki67, Cohort) %>%
  pivot_longer(-Cohort, names_to = "Variable", values_to = "Value") %>%
  mutate(Variable = factor(Variable, levels = c("log_PFS", "age", "bmi",
                                                 "tumor_size", "lymph_nodes", "Ki67"))) %>%
  ggplot(aes(x = Value, fill = Cohort)) +
  geom_density(alpha = 0.4) +
  facet_wrap(~ Variable, scales = "free", nrow = 2) +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")
print(p)
```

```{r cat-compare, fig.cap="Categorical variable distributions across cohorts.", fig.width=7, fig.height=5}
p <- combined %>%
  select(all_of(cat_vars), Cohort) %>%
  pivot_longer(-Cohort, names_to = "Variable", values_to = "Level") %>%
  count(Cohort, Variable, Level) %>%
  group_by(Cohort, Variable) %>%
  mutate(Prop = n / sum(n)) %>%
  ggplot(aes(x = factor(Level), y = Prop, fill = Cohort)) +
  geom_col(position = "dodge", alpha = 0.7) +
  facet_wrap(~ Variable, scales = "free", nrow = 2) +
  labs(x = "Level", y = "Proportion") +
  theme_bw() +
  scale_fill_brewer(palette = "Set1")
print(p)
```

```{r cohort-summary}
comp_A <- A %>%
  select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS) %>%
  summarise(across(everything(), ~ paste0(round(mean(.x), 2), " (", round(sd(.x), 2), ")"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Cohort A (1990-1995)")

comp_B <- B %>%
  select(age, bmi, tumor_size, lymph_nodes, Ki67, log_PFS) %>%
  summarise(across(everything(), ~ paste0(round(mean(.x), 2), " (", round(sd(.x), 2), ")"))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Cohort B (2000-2005)")

kable(left_join(comp_A, comp_B, by = "Variable"),
      caption = "Mean (SD) comparison of continuous variables between cohorts.")
```

## Build New Model for datB

```{r model-B-setup}
# 80/20 train-test split for datB
set.seed(8106)
trainIndex_B <- createDataPartition(B$log_PFS, p = 0.8, list = FALSE)
train_B <- B[trainIndex_B, ]
test_B <- B[-trainIndex_B, ]

x_train_B <- model.matrix(log_PFS ~ ., data = train_B)[, -1]
y_train_B <- train_B$log_PFS
x_test_B <- model.matrix(log_PFS ~ ., data = test_B)[, -1]
y_test_B <- test_B$log_PFS

cv_folds_B <- createFolds(y_train_B, k = 10, returnTrain = TRUE)
ctrl_best_B <- trainControl(method = "cv", number = 10, index = cv_folds_B,
                            selectionFunction = "best")
ctrl_1se_B  <- trainControl(method = "cv", number = 10, index = cv_folds_B,
                            selectionFunction = "oneSE")
```

```{r model-B}
set.seed(8106)
# Step 1: best on full alpha x lambda grid
enet_fit_B <- train(x_train_B, y_train_B, method = "glmnet",
                    tuneGrid = enet_grid, trControl = ctrl_best_B)

# Step 2: fix best alpha, apply 1SE along lambda only
enet_grid_1se_B <- expand.grid(alpha = enet_fit_B$bestTune$alpha,
                               lambda = exp(seq(2, -6, length = 100)))
enet_fit_B_1se <- train(x_train_B, y_train_B, method = "glmnet",
                        tuneGrid = enet_grid_1se_B, trControl = ctrl_1se_B)

pls_fit_B     <- train(x_train_B, y_train_B, method = "pls",
                       tuneGrid = data.frame(ncomp = 1:15),
                       trControl = ctrl_best_B, preProcess = c("center", "scale"))
pls_fit_B_1se <- train(x_train_B, y_train_B, method = "pls",
                       tuneGrid = data.frame(ncomp = 1:15),
                       trControl = ctrl_1se_B, preProcess = c("center", "scale"))

mars_fit_B     <- train(x_train_B, y_train_B, method = "earth",
                        tuneGrid = mars_grid, trControl = ctrl_best_B)
mars_fit_B_1se <- train(x_train_B, y_train_B, method = "earth",
                        tuneGrid = mars_grid, trControl = ctrl_1se_B)

# GAM with manual CV on datB
gam_cv_B <- sapply(seq_along(cv_folds_B), function(i) {
  train_idx <- cv_folds_B[[i]]
  gam_mod <- gam(gam_formula, data = train_B[train_idx, ])
  pred <- predict(gam_mod, newdata = train_B[-train_idx, ])
  RMSE(pred, y_train_B[-train_idx])
})
gam_fit_B_full <- gam(gam_formula, data = train_B)
```

```{r model-compare-B}
caret_models_B <- list(
  "ElasticNet (best)" = enet_fit_B,     "ElasticNet (1se)" = enet_fit_B_1se,
  "PLS (best)"        = pls_fit_B,      "PLS (1se)"        = pls_fit_B_1se,
  "MARS (best)"       = mars_fit_B,     "MARS (1se)"       = mars_fit_B_1se
)

cv_results_B <- sapply(caret_models_B, function(m) merge(m$bestTune, m$results)$RMSE)
test_preds_B <- lapply(caret_models_B, function(m) predict(m, newdata = x_test_B))
test_rmse_all_B <- sapply(test_preds_B, function(p) RMSE(p, y_test_B))
n_vars_B <- sapply(caret_models_B, count_vars)

gam_test_rmse_B <- RMSE(predict(gam_fit_B_full, newdata = test_B), y_test_B)

comp_B_table <- rbind(
  data.frame(Model = names(caret_models_B),
             CV_RMSE = round(cv_results_B, 4),
             Test_RMSE = round(test_rmse_all_B, 4),
             Num_Predictors = n_vars_B),
  data.frame(Model = "GAM",
             CV_RMSE = round(mean(gam_cv_B), 4),
             Test_RMSE = round(gam_test_rmse_B, 4),
             Num_Predictors = 15)
)
kable(comp_B_table, row.names = FALSE,
      caption = "Model performance with best and 1SE selection rules (datB).")

# For downstream use
test_rmse_B <- c(
  ElasticNet = unname(test_rmse_all_B["ElasticNet (best)"]),
  PLS        = unname(test_rmse_all_B["PLS (best)"]),
  MARS       = unname(test_rmse_all_B["MARS (best)"]),
  GAM        = gam_test_rmse_B
)

resamp_B <- resamples(list(
  ElasticNet = enet_fit_B, PLS = pls_fit_B, MARS = mars_fit_B
))
rmse_all_B <- c(
  sapply(caret_models_B[c(1,3,5)], function(m) merge(m$bestTune, m$results)$RMSE),
  GAM = mean(gam_cv_B)
)
names(rmse_all_B) <- c("ElasticNet", "PLS", "MARS", "GAM")

r2_caret_B <- resamp_B$values %>%
  select(ends_with("~Rsquared")) %>%
  colMeans(na.rm = TRUE)
names(r2_caret_B) <- gsub("~Rsquared", "", names(r2_caret_B))

gam_r2_B <- sapply(seq_along(cv_folds_B), function(i) {
  train_idx <- cv_folds_B[[i]]
  gam_mod <- gam(gam_formula, data = train_B[train_idx, ])
  pred <- predict(gam_mod, newdata = train_B[-train_idx, ])
  caret::R2(pred, y_train_B[-train_idx])
})
r2_all_B <- c(r2_caret_B, GAM = mean(gam_r2_B))
```

### Variable Importance Comparison

```{r vip-compare, fig.cap="Variable importance comparison (MARS) between cohorts.", fig.width=7, fig.height=3.5}
p1 <- vip(mars_fit, num_features = 10) + theme_bw() + ggtitle("MARS (best) - Cohort A")
p2 <- vip(mars_fit_B_1se, num_features = 10) + theme_bw() + ggtitle("MARS (1se) - Cohort B")
grid.arrange(p1, p2, ncol = 2)
```

### Coefficient Comparison (Elastic Net)

```{r coef-compare}
coef_A <- coef(enet_fit$finalModel, s = enet_fit$bestTune$lambda)
coef_Bnew <- coef(enet_fit_B$finalModel, s = enet_fit_B$bestTune$lambda)

coef_comp <- data.frame(
  Variable = rownames(coef_A),
  Cohort_A = round(as.numeric(coef_A), 4),
  Cohort_B = round(as.numeric(coef_Bnew), 4)
) %>%
  mutate(Change = round(Cohort_B - Cohort_A, 4)) %>%
  filter(Cohort_A != 0 | Cohort_B != 0) %>%
  arrange(desc(abs(Change)))

kable(coef_comp, caption = "Elastic net coefficient comparison between cohorts.",
      row.names = FALSE)
```

### MARS Comparison

```{r mars-B-summary}
summary(mars_fit_B_1se$finalModel)
```

## Summary

```{r final-summary}
best_B_name <- "MARS (1se)"

# MARS 1se CV RMSE for datB
mars_1se_cv_B <- merge(mars_fit_B_1se$bestTune, mars_fit_B_1se$results)$RMSE
mars_1se_test_B <- unname(test_rmse_all_B["MARS (1se)"])

final_table <- data.frame(
  Metric = c("Best Model", "Selection", "CV RMSE", "Test RMSE", "Num Predictors"),
  Cohort_A = c("MARS", "best",
               round(rmse_all["MARS"], 4),
               round(test_rmse["MARS"], 4),
               6),
  Cohort_B = c("MARS", "1se",
               round(mars_1se_cv_B, 4),
               round(mars_1se_test_B, 4),
               6)
)
kable(final_table, caption = "Final model summary for each cohort.", row.names = FALSE)
```

```{r save-objects, include=FALSE}
save.image("models.RData")
```
